---
title: "Class 11 Lab Session"
author: "Yvonne Yu A16333006"
format: pdf
---

### Section 4 of the PDF: Population Scale Analysis

> Q13. Read this file into R and determine the sample size for each genotype and their 
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

The following shows a breakdown of the sample size for each genotype. 

```{r}
table(expr$geno)
```

Pulls the package that would be used. 

```{r}
library(dplyr)
```
The following code gives the summary information for each of the genotypes, which can be used to identify the median. 
```{r}

# The function takes the target input genotype and prints out the genotype
# and the associated summary
geno_info <- function(gen) {
  
  # prints out the genotype inputted
  print(gen)
  
  # filters out the dataframe to only the sample size with that genotype
  new_expr <- expr %>% filter(geno == gen) 
  
  # prints out the summary of the dataframe 
  print(summary(new_expr))
}

# the list of the genotypes in the dataframe
geno_list <- unique(expr$geno)

# test run
for(i in geno_list) {
  geno_info(i)
}
```

> Q14.Generate a boxplot with a box per genotype, what could you infer from the relative 
expression value between A/A and G/G displayed in this plot? Does the SNP effect the 
expression of ORMDL3?

Pulls the package that would be utilized for this question. 

```{r}
library(ggplot2)
```


The following code is used to generate a boxplot that shows the distribution of expression for each of the genotypes. 
```{r}
ggplot(expr, aes(geno, exp, fill = geno)) + geom_boxplot(notch = TRUE)
```

Based on the boxplot, it can be inferred that the A/A genotype is expressed more, followed by the A/G genotype, and the G/G genotype being the least expressed of the three. Therefore, this would mean that the SNP genotype can affect the ORMDL3 expression levels. 

