---
title: "Homework 6"
author: "Yvonne Yu A16333006"
format: pdf
---

The following code declares the library packages that are needed to be used. Installation of the package was conducted through `install.packages("bio3d")`.
```{r}
library(bio3d)
```

The following code gives a breakdown of the proposed function. 

The function is aimed to take a specified protein of interest and generate a plot based on the protein structure of chain A and the targeted column of focus. 
```{r}
#the input of the function is the specified protein of interest
newfun <- function(input) {

    #reads the input protein and assigns it to variable s
    s <- read.pdb(input)

    #isolates the protein structure to the targeted chain
    s.chainA <- trim.pdb(s, chain="A", elety="CA")

    #assigns a variable to the targeted column of focus
    s.b <- s.chainA$atom$b

    #plots the vector, adding the secondary structure, 
    #assigning the type of plot and the title of the y-axis
    plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor")
    }
```

The following code runs the function to make sure it is successful. 
```{r}
#the sample list of variables to be inputted
protein_list <- c("4AKE", "1AKE", "1E4Y")

#run
for (i in protein_list) {
    print(i)
    newfun(i)
    }

```


